Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-06 paper.tex
Light edits to intro

\usepackage[english]{babel}
\usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

\usepackage{authblk}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{url}
% JK: turning this off for the moment as I keep clicking through on links
% to the bibliography while reading the text and it's intensely annoying.
% Can reinstate when we're ready to preprint
% \usepackage[hidelinks]{hyperref}

@@

Recombination via template switching is a common feature in the evolution of
coronaviruses (CoV) \citep{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genomic diversity, with recombinant lineages
associated with CoV adaptation to new host species and with the production of
more immune evasive variants \citep{Graham2010-xe,De_Klerk2022-tt}.

Recombination was challenging to detect in SARS-CoV-2 early in the pandemic due
to gradual accumulation of genetic variation. In late 2020, the appearance and
spread of variants of concern (VoC) harbouring multiple characteristic
mutations increased the detectability of recombination. By searching for
genomes containing segments from different VoC, multiple recombinant lineages
have been detected, for example, between Alpha and pre-VoC
\citep{Jackson2021-ik}, Alpha and Epsilon \citep{Wertheim2022-hj}, and Alpha and
Delta \citep{Sekizuka2022-xz}). More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for recombination
\citep{Bal2022-hq}. Recombinant lineages that have sufficient evidence of onward
transmission (more than five descendants) and monophyly (> 70\% bootstrap
support) are eligible for a Pango lineage designation \citep{Rambaut2020-dw},
starting with the letter X (“Pango Lineage Nomenclature: Provisional Rules for
Naming Recombinant Lineages” 2021).

Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
combine specific subsets of mutations from different lineages (e.g.,
\citep{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}).
This approach only identifies those inter-lineage recombination events
resulting in the patterns investigated. Others have inferred recombinant
genealogies to disentangle recombination from recurrent mutation, without
having to pre-specify characteristic mutations for each lineage
\citep{Ignatieva2021-rg}. More recently, Turakhia et al. \citep{Turakhia2022-it}
has presented a phylogeny-based approach ("RIPPLES") to identify putative
recombinants among over ten million SARS-CoV-2 genomes, without pre-specifying
motifs. To find candidate recombinants, RIPPLES first builds a phylogeny
without assuming recombination and then scans for long branches (i.e., carrying
many mutations) and then determines if these candidates would be better
explained by recombination. Specifically, RIPPLES exhaustively breaks each
candidate sequence into segments and attempts to find more parsimonious
placements for each segment on the phylogeny. If such placements are found, the
sequence is identified as a putative recombinant. Although it enables rapid
searches for genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2
phylogeny that accounts for only mutations, treats recombinants \textit{post
hoc}, and is an incomplete representation of the reticulate evolutionary
history of SARS-CoV-2. As noted by the authors, a \textit{post hoc} treatment
of recombination is possible when recombinant lineages are rare and leave few
descendants, but this is becoming increasingly untenable with the proliferation
of recombinant lineages (e.g., more than half of the sequences sampled in
February, 2023 are from XBB and its descendants; \citep{Chen2022-pz}.

Inferring evolutionary trees while accounting for both mutation and
recombination is a long-standing challenge. This challenge is particularly
acute for SARS-CoV-2, given the large amount of data that has been shared, with
over 15 million sequences in the GISAID database \citep{Shu2017-hp}. It is well
known that recombination distorts phylogenies \citep{Schierup2000-fg} and
affects the results of downstream analysis (e.g., inference of selection;
\citep{Anisimova2003-vr}). However, standard phylogenetic methods do not account
for recombination (e.g., \citep{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}).
Alternatively, genetic ancestries of recombining lineages can be represented by
ancestral recombination graphs (ARG; \citep{Griffiths1981-lw,Hudson1983-if},
which  capture network-like phylogenetic relationships \citep{Gusfield2014-qw}.

Approaches to infer ARGs now exist that can scale to tens of thousands of human
genomes and beyond
\citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}. Recently,
Wohns et al. \citep{Wohns2022-th} reconstructed a unified whole-genome genealogy
of thousands of modern and ancient human genomes, based on over 90 million
variable sites. To achieve this, Wohns et al. \citep{Wohns2022-th} utilised
several computational advances in ARG inference: (1) the "succinct tree
sequence" data structure that can compactly store large ARGs
\citep{Kelleher2018-xc}; (2) a highly efficient Hidden Markov Model (HMM)
implementation of the Li and Stephens (LS) model (Li and Stephens 2003) to
infer copying paths between genomes \citep{Kelleher2019-ba}; and (3) efficient,
open-source software tools
\citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

Here, we present a preliminary version of a novel scalable methodology
(“sc2ts”, pronounced “scoots”) that builds on the advances above and
additionally takes advantage of known collection dates to infer ARGs from
millions of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2
genome sequences, we demonstrate that sc2ts can be used to simultaneously (1)
detect putative SARS-CoV-2 recombinants; (2) infer the phylogenetic
relationships of the major SARS-CoV-2 lineages; (3) reconstruct the mutational
spectrum of SARS-CoV-2; and (4) provide a compact record of evolutionary
information about SARS-CoV-2 that can be efficiently accessed using "tskit"
(tree sequence toolkit) software library \citep{Kelleher2018-xc}. Sc2ts is a
promising method to automate detection of SARS-CoV-2 recombinants and to
jointly model recombinational and mutational processes shaping the genomic
diversity of SARS-CoV-2 at large scale.

>>>>

\usepackage[round]{natbib}
\usepackage[english]{babel}
\usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

\usepackage{authblk}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{url}
% JK: turning this off for the moment as I keep clicking through on links
% to the bibliography while reading the text and it's intensely annoying.
% Can reinstate when we're ready to preprint
% \usepackage[hidelinks]{hyperref}

@@

Recombination via template switching is a common feature in the evolution of
coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genomic diversity, with recombinant lineages
associated with adaptation to new host species and with the production of
more immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.

Recombination was challenging to detect in SARS-CoV-2 early in the pandemic due
to gradual accumulation of genetic variation. In late 2020, the appearance and
spread of variants of concern (VoC) harbouring multiple characteristic
mutations increased the detectability of recombination. By searching for
genomes containing segments from different VoC, multiple recombinant lineages
have been detected, for example, between Alpha and pre-VoC
\citep{Jackson2021-ik}, Alpha and Epsilon \citep{Wertheim2022-hj}, and Alpha and
Delta~\citep{Sekizuka2022-xz}. More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for recombination
\citep{Bal2022-hq}. Recombinant lineages that have sufficient evidence of onward
transmission (more than five descendants) and monophyly
($>70\%$ bootstrap support) are eligible for a Pango lineage
designation~\citep{Rambaut2020-dw},
starting with the letter X
% FIXME: should this be a citation?
(``Pango Lineage Nomenclature: Provisional Rules for Naming Recombinant Lineages'' 2021).

Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
combine specific subsets of mutations from different lineages
\citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}.
This approach only identifies those inter-lineage recombination events
resulting in the patterns investigated. Others have inferred recombinant
genealogies to disentangle recombination from recurrent mutation, without
having to pre-specify characteristic mutations for each lineage
\citep{Ignatieva2021-rg}. More recently, \cite{Turakhia2022-it}
presented a phylogeny-based approach (``RIPPLES'') to identify putative
recombinants among over ten million SARS-CoV-2 genomes, without pre-specifying motifs.
To find candidate recombinants, RIPPLES first builds a phylogeny
without assuming recombination and then scans for long branches (i.e., carrying
many mutations) and then determines if these candidates would be better
explained by recombination. Specifically, RIPPLES exhaustively breaks each
candidate sequence into segments and attempts to find more parsimonious
placements for each segment on the phylogeny. If such placements are found, the
sequence is identified as a putative recombinant. Although it enables rapid
searches for genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2
phylogeny that accounts for only mutations, treats recombinants
\textit{post hoc}, and is an incomplete representation of the reticulate evolutionary
history of SARS-CoV-2. As noted by the authors, a \textit{post hoc} treatment
of recombination is possible when recombinant lineages are rare and leave few
descendants. However, this is becoming increasingly untenable with the proliferation
of recombinant lineages; for example, more than half of the sequences sampled in
February 2023 are from XBB and its descendants~\citep{Chen2022-pz}.

Inferring evolutionary trees while accounting for both mutation and
recombination is a long-standing challenge. This challenge is particularly
acute for SARS-CoV-2, given the large amount of data that has been shared, with
over 15 million sequences in the GISAID database \citep{Shu2017-hp}. It is well
known that recombination distorts phylogenies \citep{Schierup2000-fg} and
affects the results of downstream analysis, e.g., inference of
selection~\citep{Anisimova2003-vr}.
However, standard phylogenetic methods do not account
for recombination \citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}.
Alternatively, genetic ancestries of recombining lineages can be represented by
ancestral recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-if},
which  capture network-like phylogenetic relationships \citep{Gusfield2014-qw}.

Approaches to infer ARGs now exist that can scale to tens of thousands of human
genomes and beyond
\citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}. Recently,
\cite{Wohns2022-th} reconstructed a unified whole-genome genealogy
of thousands of modern and ancient human genomes, based on over 90 million
variable sites. To achieve this, \cite{Wohns2022-th} utilised
several computational advances in ARG inference: (1) the "succinct tree
sequence" data structure that can compactly store large ARGs
\citep{Kelleher2018-xc}; (2) a highly efficient Hidden Markov Model (HMM)
implementation of the Li and Stephens (LS) model (Li and Stephens 2003) to
infer copying paths between genomes \citep{Kelleher2019-ba}; and (3) efficient,
open-source software tools
\citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

Here, we present a preliminary version of a novel scalable methodology
(``sc2ts'', optionally pronounced ``scoots'') that builds on the advances above and
additionally takes advantage of known collection dates to infer ARGs from
millions of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2
genome sequences, we demonstrate that sc2ts can be used to simultaneously (1)
detect putative SARS-CoV-2 recombinants; (2) infer the phylogenetic
relationships of the major SARS-CoV-2 lineages; (3) reconstruct the mutational
spectrum of SARS-CoV-2; and (4) provide a compact record of evolutionary
information about SARS-CoV-2 that can be efficiently accessed using "tskit"
(tree sequence toolkit) software library \citep{Kelleher2018-xc}. Sc2ts is a
promising method to automate detection of SARS-CoV-2 recombinants and to
jointly model recombinational and mutational processes shaping the genomic
diversity of SARS-CoV-2 at large scale.
